---
title: "SIGNAL integration for unlabeled datasets"
author: "Yang Zhou"
date: "2024-12-01"
output: html_document
package: SIGNAL
vignette: |
  %\VignetteIndexEntry{SIGNAL: exploring group-specific technical variation patterns of single-cell data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

We can use SIGNAL to perform multi-scale analysis of single-cell data to identify cell subtypes specific to tissues, conditions, and developmental stages.

```{r, message=FALSE, warning=FALSE}
library(SIGNAL)
library(irlba)
library(uwot)
library(ggpubr)
library(randomcoloR)
library(cowplot)
set.seed(2024)
```

# 2. Application on the developing human immune cell atlas

In this vignette, we first demonstrate on a recently published [developing human immune cell atlas](https://www.science.org/doi/10.1126/science.abo0510) how SIGNAL can be used to integrate in a way that preserves differences between tissues and developmental stages.

# 2.1 Load data matrix and metadata

We perform data integration on the normalized scRNA-seq gene expression matrix. The HVGs provided by the authors are used.

```{r, message=TRUE, warning=FALSE}
X = readRDS("/home/server/zy/group_scripts/datasets_preparation/Developing_immune/X.rds")
meta = readRDS("/home/server/zy/group_scripts/datasets_preparation/Developing_immune/meta.rds")
str(X)
str(meta)
```

# 2.2 Visualization of raw data

```{r, message=TRUE, warning=FALSE, fig.width=19, fig.height=5}
Colors = distinctColorPalette(13)
pca_res = irlba(t(X), nv = 50)
raw_emb = as.matrix(pca_res$u %*% diag(pca_res$d))
raw_umap = as.data.frame(umap(raw_emb))
colnames(raw_umap) = c("UMAP1", "UMAP2")
raw_umap = cbind.data.frame(meta, raw_umap)
p1 = ggscatter(raw_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "CellType", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
p2 = ggscatter(raw_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Tissue", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
p3 = ggscatter(raw_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Stage", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
plot_grid(p1, p2, p3, align = 'h', axis = "b", nrow = 1)
```

# 2.3 SIGNAL integration using 'tissue' and 'stage' as group variables in a two-group setting

```{r, message=TRUE, warning=FALSE, fig.width=19, fig.height=5}
signal_emb = Run.gcPCA(X, meta, g_factor = c("Tissue", "Stage"), b_factor = "Batch")
signal_umap = as.data.frame(umap(t(signal_emb)))
colnames(signal_umap) = c("UMAP1", "UMAP2")
signal_umap = cbind.data.frame(meta, signal_umap)
q1 = ggscatter(signal_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "CellType", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
q2 = ggscatter(signal_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Tissue", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
q3 = ggscatter(signal_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Stage", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
plot_grid(q1, q2, q3, align = 'h', axis = "b", nrow = 1)
```

# 3. Application on the mouse islet atlas

We next apply SIGNAL to integrate a recently published [mouse islet atlas](https://doi.org/10.1038/s42255-023-00876-x) using a three-group setting.

# 3.1 Load data matrix and metadata

We perform data integration on the normalized scRNA-seq gene expression matrix. The HVGs provided by the authors are used.

```{r, message=TRUE, warning=FALSE}
X = readRDS('/home/server/zy/group_scripts/datasets_preparation/Mouse_islet_atlas/X.rds')
meta = readRDS('/home/server/zy/group_scripts/datasets_preparation/Mouse_islet_atlas/Meta.rds')
str(X)
str(meta)
```

# 3.2 Visualization of raw data

```{r, message=TRUE, warning=FALSE, fig.width=19, fig.height=14}
Colors = distinctColorPalette(11)
pca_res = irlba(t(X), nv = 50)
raw_emb = as.matrix(pca_res$u %*% diag(pca_res$d))
raw_umap = as.data.frame(umap(raw_emb))
colnames(raw_umap) = c("UMAP1", "UMAP2")
raw_umap = cbind.data.frame(meta, raw_umap)
p1 = ggscatter(raw_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "CellType", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
p2 = ggscatter(raw_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Sex", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
p3 = ggscatter(raw_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Disease", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
p4 = ggscatter(raw_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Stage", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
plot_grid(p1, p2, p3, p4, align = 'h', axis = "b", nrow = 2)
```

# 3.3 SIGNAL integration using 'sex', 'disease', and 'stage' as group variables in a two-group setting

```{r, message=TRUE, warning=FALSE, fig.width=19, fig.height=14}
signal_emb = Run.gcPCA(X, meta, g_factor = c("Sex", "Disease", "Stage"), b_factor = "Batch")
signal_umap = as.data.frame(umap(t(signal_emb)))
colnames(signal_umap) = c("UMAP1", "UMAP2")
signal_umap = cbind.data.frame(meta, signal_umap)
q1 = ggscatter(signal_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "CellType", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
q2 = ggscatter(signal_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Sex", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
q3 = ggscatter(signal_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Disease", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
q4 = ggscatter(signal_umap, x = "UMAP1", y = "UMAP2", size = 0.1, color = "Stage", palette = Colors, legend = "right") + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
plot_grid(q1, q2, q3, q4, align = 'h', axis = "b", nrow = 2)
```

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>
